\documentclass[a4paper,12pt]{article} 

\usepackage[spanish]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{geometry}
\geometry{margin=2.5cm}

\usepackage{hyperref}
\usepackage{microtype} 

\usepackage{fancyhdr}

\pagestyle{fancy}
\fancyhf{}
\lhead{Class Notes}
\rhead{The Culture of Business}
\cfoot{\thepage}

\title{\vspace{-2cm}\textbf{Class Notes}\\[0.2cm]
\large The Culture of Business}
\author{Juan Ignacio Elosegui}
\date{Universidad Torcuato Di Tella \\ Marzo 2026}

\begin{document}
\maketitle
\section*{First Part}
\subsection*{Class I - 03.03}
\noindent \textbf{What is ``The Culture of Business, Finance, and the Economy''?} \\
\noindent This course examines the \textbf{practices, intuitions, and less-examined kinds of knowledge} that make business, finance, and the economy function in a particular way. It draws on case studies ranging from Norwegian management consultants and Swiss investment bankers to medieval traders and Native North American shamans. \\
\\
\noindent \textbf{Objective:} \\
Examine the ways in which business, finance and the economy work, think, and make sense of things \emph{beyond the claims these professions and disciplines make about themselves}. The aim is not to antagonise or ``debunk'' them, but rather to reflect on:
\begin{itemize}
	\item Why certain ways of doing things stuck instead of others;
	\item How specific ideas came to mean what they do and with what consequences;
	\item How certain technologies and ways of knowing can shape the things they purport to merely describe.
\end{itemize}

\noindent This way of thinking and of doing things is in itself a \textbf{culture} that requires as much learning, nurturing, reproducing and ritualising as any other culture does. \\
\\
\noindent \textbf{What is culture?}
\begin{itemize}
	\item Culture is a \textbf{way of seeing things}, a \textbf{way of thinking}, a \textbf{way of making sense}.
	\item Culture influences how we experience things, how we think about, apprehend and make sense of complex notions like normality and deviance, the good and the bad, correctness and impropriety; how we ``sort through'' what matters and does not matter.
	\item Cultures are \textbf{learned and mediated by artefacts, technologies and knowledge} that reproduce it, sustain it, and make it. \\
\end{itemize}

\noindent \textbf{Technologies and ways of knowing in TCBFE:} \\
All of these technologies, artefacts and ways of knowing create:
\begin{itemize}
	\item A way of making sense of the world;
	\item A way of ordering the things we encounter;
	\item A way of deciding what matters and what not, when, and for what;
	\item A way of persuading ourselves and trying to persuade others;
	\item A way of generating a common sense and an intuition. \\
\end{itemize}
\subsection*{Class II - 05.03}
\noindent According to the reading \textit{The Superiority of Economics}, economics is \textbf{seen as superior} not only intellectually but also professionally, politically, economically and morally. \\
This superiority is \textbf{linked to strong reliance on formal, mathematical methods; high market demand, tight disciplinary control, and high compensation}. \\
\\
\noindent But, why is it \emph{superior}?
\begin{itemize}
	\item \textbf{Insularity.} Economics is highly \emph{self-referential}: most citations are within economics itself, and economists are \emph{not likely to engage with other disciplines}.
	\item \textbf{Method.} They justify their superiority through \emph{mathematical models, abstraction and simplification}. This leads to \emph{viewing other sciences as less rigorous} and expand to other fields using tools such as game theory, rational choice theory and general equilibrium models.
	\item \textbf{Disciplinary Control.} Here is a tight control over what counts as ``proper economics''. Academic departments, top journals, and PhD programs enforce a \emph{unified orthodoxy}, filtering out heterodox approaches (Marxist economics).
	\item \textbf{Market Demand \& Prestige.} Economists hold \emph{key} positions in governments, central banks, and international organizations. Their expertise is highly valued in the private sector (consulting), granting them \emph{significant political and economic power}.
	\item \textbf{Moral Authority.} Economics presents itself as a ``science'' that can objectively determine the most efficient outcomes, often \emph{framing policy choices as technical necessities rather than political decisions}, which gives it a \emph{moral high ground}.
\end{itemize}
\subsection*{Class III - 10.03 (Economists as Shamans)}
\noindent \textbf{What is a shaman?} \\
\noindent Shamans mediate between worlds. They act on behalf of the community, or of individuals, healing, harming, translating and interpreting knowledge, mediating and producing obscure, secret or forbidden knowledge. They translate the unseen, engaging the gods, the world, nature and anything beyond ``normal'' human abilities through techniques and knowledge only they have, artefacts only they are allowed or effectively able to use. \\
\\
\noindent \textbf{Rites of Passage} \\
\noindent They are the culturally approved and culturally understood moments where one's status changes: one becomes something else. These moments test us, push us to a certain limit, sometimes defile or degrade us, to mark us as able to handle, produce, or mediate certain kinds of knowledge or practices. To make it past them is to become worthy of that which is kept beyond that test. \\
\\
\noindent \textbf{Example: Investment banking interns}
\begin{itemize}
	\item Highly competitive environment: fighting 1500 candidates for one of two places.
	\item Brutal hours: working through the night is almost a \textbf{rite of passage}.
	\item The ``magic roundabout'': taxi home at 5AM, shower, put on a fresh shirt, back to your desk.
	\item Rules against overwork are impossible to implement because junior staff will work around policies to prove themselves.
	\item As a human resources consultant put it: ``Stating that one should not work on Saturdays is like telling people in a bar that they should drink less. In other words, there is a culture and there are rules. If these rules do not reflect the culture, rules are useless.''
\end{itemize}

\noindent \textbf{Why does this matter?} \\
\noindent The market as an order — not whether ``it works'' or not, but what matters is the \emph{kind of sense it makes} when we think about it in these terms. The culture of economics involve:
\begin{itemize}
	\item A particular way of understanding order.
	\item A particular \emph{figure of authority} (the economist as shaman, ``one who knows'').
	\item Particular rituals and tests (rites of passage that \emph{produce legitimate knowers}).
\end{itemize}
\subsection*{Class IV - 12.03}
\noindent \textbf{Mathematics: from divine perfection to a deal with Satan.} \\
\noindent How did mathematics become what it is today, and how did it become part of economics? Was it always ``clear'', ``objective'', ``perfect''?  Did we discover it or did we invent it as a universal means of knowledge? \\
\\
\noindent \textbf{Mathematics in the 1600s: Bacon vs. words} \\
\noindent In 1623, Francis Bacon argued that words ``do violence to the understanding'' and that we needed ``a new and deeper remedy.'' The question was: where was that new remedy? \\
\\
\noindent \textbf{What we call mathematics did not exist back then} \\
\noindent The heart of mathematics was \emph{geometry}. When Galileo stated that God wrote the book of the world in the language of mathematics, he was referring to \emph{geometric diagrams}, not in symbolic notation. The Seven Liberal Arts \emph{did not include} ``mathematics'' in the modern sense. \\
\noindent Descartes held that Arithmetic and Geometry were \emph{free from any taint of falsity or uncertainty} because they derived knowledge through deduction, not from experience. \\
\\
\noindent \textbf{The limits of Geometry} \\
\noindent In Geometry, quantities (``four apples'') and magnitudes (four square meters) were different \emph{kinds} of numbers. So the equation: \( ax^2 + bx + c = 0 \) did not even make sense, because it ``pretends'' to add linear and square kinds. \\
\noindent \textbf{Alogos.} Since Euclid and the beginnings of Geometry, everyone knew of lines whose lengths cannot be expressed as multiples of a common unit (e.g., $\sqrt{2}$). These irrational numbers posed a \emph{deep conceptual problem}. \\
\\
\noindent \textbf{Proto-algebra: a plebeian practice} \\
\noindent A \emph{proto-algebra} existed, but it was considered dirty, plebeian, and \emph{not real mathematics}. It was perceived to lack rigorous standards of proof. Algebra was a practical, \emph{earthly kind of knowledge associated with common activities}: merchandising, sailing, and trade-related calculations. \\
\noindent \textbf{Al-Khw\=arizm\=\i} wrote one of the foundational texts of algebra in 820 AD. The word \emph{algebra} itself derives from \textit{al-jabr}, usually translated as ``completion.'' His proto-algebra was written describing step by step how to carry and place digits, use zeros, and perform operations. \\
\noindent Al-Khw\=arizm\=\i{} knew about irrational numbers such as $\sqrt{2}$; he instructed the reader that one must be content with ``an approximation, and not the exact truth: for God alone knows what the exact root is.'' \\
\noindent This proto-algebra spread across Europe through \emph{merchants}. It was primarily the business of ``mathematical practitioners'': people who taught mathematics outside of the university system through textbooks and tutoring. \\
\\
\textbf{Conflicting ideas about numbers} \\
\noindent Algebra brought with it \emph{conflicting ideas} about what numbers are and how they work. \textbf{Stifel} wrote en 1544 that it was disputed whether irrational numbers were true numbers or fictions. In their favor was their utility in calculation; against them was the fact that when we try to express them as rational numbers. \\
\noindent A major shift in the conception of number occurred: \emph{numbers began to be imagined not through the perfection of arithmetic and geometry, but through the digits themselves} --- the tools of the commoners who practiced algebra. The claim emerged that ``there are no absurd, irrational, irregular, inexplicable numbers'': one cannot write $\sqrt{2}$ as a fraction, but one can write it as $1.4142\ldots$ and continue the expansion as far as one likes. \\
\noindent Even further: cubic equations yielded roots of negative numbers. $\sqrt{-1}$ cannot even be approximated by decimal fractions. For centuries this was rejected, but the fact that algebra's procedures could produce such results seemed to imply that the procedures themselves were wrong. \\
\\
\textbf{Algebra enters mathematics}\\
\noindent \textbf{Fran\c{c}ois Vi\`ete} made a key innovation: instead of working on single equations , one could now draw conclusions about general ``species'' of equations. He ended his \textit{Introduction} with the statement of purpose: ``TO LEAVE NO PROBLEM UNSOLVED.'' \\
\noindent \textbf{William Oughtred} pushed this further. Symbols became a medium free of rhetoric: the naked truth, unadorned, all substance. And so mathematics \emph{became thinkable as a means to universal knowledge}. \\
\\
\textbf{At what cost? The Faustian bargain.} \\
\noindent The title of the class --- ``from divine perfection to a deal with Satan'' --- captures the tension: mathematics gained its power as a universal language by \emph{incorporating the ``dirty'', practical, merchant algebra} and by \emph{accepting entities} (irrationals, imaginaries) that violated the classical ideals of geometric perfection. As Goethe wrote: ``The alphabet is really now superfluous / for in this sign all men can find salvation.'' \\
\noindent The cost was \emph{accepting numbers and operations that did not correspond to anything in the physical world} --- a Faustian bargain for universal explanatory power (you got the power that you wanted, but at the cost of ``fake'' numbers).
\subsection*{Class V - 17.03}
\noindent \textbf{Before the 1870s.} \\
\noindent The discipline was called \textbf{political economy} and \emph{resembled morals and politics} more than mathematics. Figures like Adam Smith, and John Stuart Mill worked with principles (not laws), specific markets, practical knowledge, and non-mathematical models. Mill even declared in 1848 that the theory was ``complete'' --- nothing left to clear up. \\

\noindent \textbf{The Crisis of the 1870s.} \\
\noindent Reality contradicted that supposed ``completeness'': labor protests, strikes, growing national debt, battles over free trade. Clearly, \emph{plenty} remained unresolved.

\noindent The turning point had a name: \textbf{William Stanley Jevons}. He wanted ``to fling aside, once and for ever, the preposterous assumptions of political economy'' and \emph{rebuild the discipline} on Bentham's utilitarianism (humans as creatures who minimize pain and maximize pleasure). His central claim was: ``Economics must be mathematical, simply \emph{because it deals with quantities}''.

\noindent But, \emph{which} mathematics? The arithmetic of the Cambridge Tripos couldn't handle how tiny amounts of goods produce tiny increases in happiness. The solution came from Leibniz's calculus (the \emph{derivative}), which allowed him to develop the principle of diminishing marginal utility, the bedrock of contemporary economics. \\
\\
\noindent \textbf{The cost.} \\
\noindent Everything that didn't fit into calculus (class struggle, moral economy, land ownership debates) was expelled as ``non-economic knowledge''. Jevons \emph{freed economics from its material constraints} --- no more need to know about farms, ships, or coal. You could now do it at your desk. \\
\\
\noindent \textbf{Critical Reception.} \\
\noindent Even skilled mathematicians like Alfred Marshall were wary: he used mathematics sparingly because he believed a good mathematical theorem on economic hypotheses was \emph{unlikely} to be good economics. \\
\noindent More recently, Krugman argued economists \emph{mistook mathematical beauty for truth}, and Piketty spoke of a ``childish passion for mathematics'' at the expense of historical research. \\
\\
\noindent \textbf{Why Did It Stick?}
\begin{enumerate}
	\item \textbf{Rites of passage.} Economics students feel they must master formal modeling to compete, even if they doubt its real usefulness (\underline{Spiegler}).
	\item \textbf{Status anxiety.} Unlike engineers or doctors, economists can't point to concrete objects or predictive power; \emph{mathematics makes them ``feel smart'' and valuable}.
	\item \textbf{Authority in a technocratic world.} Mathematics in economics mostly serves to say ``Look at how very scientific I am'' (\underline{McCloskey}).
	\item \textbf{The genie has left the bottle.} Economists \emph{created economic phenomena through mathematics}, and now there's no way of knowing them except through mathematics.
	\item \textbf{A bridge between tech, management, and finance,} plus a quest for truth beyond particulars and idiosyncrasies.
\end{enumerate}
\noindent \textbf{Conclusion.} \\
\noindent Schabas notes there's no obvious reason economics should ever have adopted mathematics --- but the same could be said of physics, and today's economics is inconceivable without that mathematical anchoring.

\subsection*{Class VI - 19.03 (Economics and Physics, the science of all things)}
\noindent No economics if mathematics didn't exist. Economics wouldn't make sense. \\
\\
\noindent What does trade have to do with gravity? ``to flow'' and ``to swell'' also have to do with physics, some words used in businesses are also used in physics! Such as liquidity, flows, pipes, bubbles, and so on. There IS a reason for this. \\
\\
\noindent Mathematics is the language of physics, and physics have to do with economics as we can tell. There was an urge to create such a connection, it didn't happen naturally. Some concepts were borrowed from natural sciences. \\
\\
\noindent There was an active urge to create this connection: it \emph{didn't happen naturally}. 19th-century economists consciously \emph{borrowed concepts} from the natural sciences to \emph{justify} the use of mathematical reasoning and to ``discover'' general laws of economics. Portraying economics as a natural science was also \emph{an attempt to endow it with more authority}. ``Economics concerns itself with the interplay of a variety of forces, and for this reason it is impossible to present it without the aid of mathematics, just as is the case with astronomy, physics, mechanics and so forth'' (\underline{Gossen}). \\
\\
\noindent \textbf{But why physics? Why did economics get such an unhealthy obsession with physics?} \\
\noindent Why not astronomy, or biology? Because by the end of the 18th century and throughout the 19th, \emph{physics was the queen of sciences}. It had huge discoveries: light, wave theory, atoms, X-rays… it seemed to be \emph{approaching a unified theory of everything}. It was the closest we'd ever felt to explaining everything.
\begin{itemize}
	\item \textbf{Fisher.} He studied general equilibrium analysis and built a hydraulic mechanism to simulate equilibrium prices and quantities. He used \emph{fluids as money}, giving flesh to his ideas with mere water and balances. He \emph{explained economics through physical phenomena and mechanisms}. He also used a mechanical balance of exchange to clarify the quantity theory of money.
	\item \textbf{Bill Phillips.} He built the MONIAC: a hydraulic machine. Separate water tanks represented households, business, government, exporting and importing sectors. Coloured water pumped around the system measured income, spending and GDP.
	\item \textbf{Walras.} In what he called ``a new branch of mathematics'', he linked economics to Newton's laws. He \emph{argued economics was a mathematical science}, exploring the analogy between speed, space and time on one side, and scarcity, utility and quantity on the other.
\end{itemize}
\noindent \textbf{Fisher's Translations.} \\
\noindent There is a direct correlation between mechanical and economic terms: particle/individual, space/commodity, energy/utility... \\
\noindent This direct correlation is still with us today. \\
\\
\noindent \textbf{Jevons's Ideas.}
\begin{enumerate}
	\item Utility is difficult even to conceive, but it is the amount of utility that prompts us to buy, sell, consume, work, produce, and we perceive it through its quantitative effects (wages, prices, etc.), just as we can no more know or measure gravity in itself, but we measure the motion of a pendulum.
	\item In practice, no market ever long fulfils the theoretical conditions of equilibrium, but every branch of physics involves abstracting from the complexities of the actual world. \emph{So we don't need to know everything in the market to believe in The Market}.
\end{enumerate}
\noindent \textbf{Is this purely our imagination?} \\
\noindent Were economics and physics ever really similar, or have we imagined them and designed them such that now our economic intuition — our economic culture — cannot imagine itself out of it?
\subsection*{Class VII - 26.03}
\noindent \textbf{What is a model in business, finance and the economy?} \\
\noindent A model is not a job division, a blueprint for building a boat, nor an early detailed map. The Tableau Économique (François Quesnay, late 1750s) is a model --- and to this day, we don't fully understand how it works. It was an economic chart designed to show King Louis XV how money circulated through his kingdom, depicting classes of people with zig-zag lines representing the flow of goods and money between them.

Why is it a model? \emph{Because you can work with it}. It's not about correct scale or fine detail; you could multiply and divide across the tableau to figure out how money fluctuated, map out different scenarios, and formulate hypotheses — exactly what models do today. \\

\noindent A model should follow the following properties:
\begin{itemize}
	\item It must be \emph{schematic}.
	\item It is never just a portrait, and it never works through analogy.
	\item It cannot be tweaked in infinite ways: it is a \emph{managed manipulation}.
	\item It must be \emph{simplified} enough to reveal what is essential and what isn't --- not so small that it becomes trivial, not so big that it becomes overly generalised.
\end{itemize}

\noindent They have two kinds of rules:
\begin{itemize}
	\item \textbf{Formal rules.} Mathematics, algebra, physics.
	\item \textbf{Substance rules.} Economic principles.
\end{itemize}

Models must have a clear endpoint. Whatever they include must be enough to know what they claim to make knowable. What's inside the model is what you need to know; anything else falls outside it. \\
\\
\noindent \textbf{How come economists fail SO much and yet we still call them?} \\
\noindent For four reasons (despite \underline{Pilmis}):
\begin{itemize}
	\item \textbf{They seek refuge in models.} The market is immense: so models focus on a subset of variables. When something unpredictable happens, economists treat it as an \emph{exogenous factor}. This unknowability becomes a resource:  errors are justified on the grounds that \emph{the market is too vast to be bounded} by any definition. Paradoxically, this leads to the conclusion that models should be improved rather than withdrawn, perpetuating reliance on them and on those who handle them.
	\item \textbf{Collective Ignorance.} It's risky to go out on a limb alone, so \emph{forecasts tend to converge}. If everyone was wrong, no individual is accountable --- the very fact that no one foresaw the event ``proves'' it was unforeseeable. Forecasts were wrong, but \emph{no error was made}.
	\item \textbf{Emphasis on narrative.} Economists present numbers \emph{alongside a whole narrative}. The figures may be wrong, but if the story matches the unfolding string of events, the forecast is considered ``right''. Forecasting becomes \emph{more about storytelling than calculation} --- you can be ``right for the wrong reasons'' and vice versa.
	\item \textbf{Emphasis on procedure and compliance.} When forecasts are way off, the last refuge is \emph{showing that the procedures were followed exactly as the book prescribed}. If the ritual of forecasting was correctly performed, the outcome cannot be criticised.
\end{itemize}
\noindent \textbf{Modeling: a belief-based practice.} \\
\noindent For all these reasons, modelling is a belief-based practice, not a result-based one. We use models because we believe in them as a way of knowing the world — they are the right way of doing things, even when they fail half the time.

This is why \underline{Pilmis} concludes that models function much like magic: because the authority of models rests on collective belief rather than evidence. The spell must be cast correctly; if the outcome fails, it's because a step was missed or because the universe (the market) holds something beyond our reach --- never because the practice itself is unfit for purpose.
\subsection*{Class VIII - 31.03}
\noindent \textbf{Thinking like an economist 101.} How economics think about certain discussions. \\
\\
\begin{itemize}
	\item Eldest people cost more to take care
	\item cost of euthanasia
	\item salvage of limbs
	\item the cost of fatigue within your employees
	\item cost of CASUAL SEX
\end{itemize}
\noindent ... so on. So many dumb shit that economists can analyze its value on a cost-benefit scenario. \\
\\
\noindent \textbf{Legal organ sales.} Will increase the supply of organs -- saving many lives. It is a hard core economic argument... how do we convince people to donate. \textbf{Getty Images case.} \\
\\
\noindent It started in the 1960s, this way of thinking expanded into non-economic topics. Environmental laws had no relation with economics, so many campaigns seeked to save the soil, the water and the air. At the time, the Clean Air Act and Clean Water Act set rules to limit pollution to certain levels. ``You cannot polute beyond this point, no matter how much you produce or who buys your product''. But economists disagreed with this measure because it didn't discriminate the producers: it is penalising those who produce ``good'' and ``bad'' stuff, (complete this part). No one cared what economists thoughts about climate change back in the 60s and 70s because some things couldn't and shouldn't be calculated: it is wrong, period. \\
\\
\noindent Both acts were revised and changed. They were changed and got some economic thinking into it: selling polluting credits to other companies. This was WAY more successful than the old acts. Economists nailed the measure.
\begin{itemize}
	\item \textbf{Feature 1.} Economic thinking organises everything in a market. Markets are the only efficient allocators of resources. Pollution is a market problem and is a failure of it -- it was nearly solved using the key terms such of choice and competition.
	\item \textbf{Feature 2.} Efficiency is everything: a good policy is KEY to reach an objective and can be evaluated through efficiency. Efficiency ignores completely the moral and political values.
\end{itemize}
\noindent (1) Politics and morality does NOT matter within the boundaries of economics, it cannot be explained with efficiency and market. (2) Economics is always finding ways to take over every single topic. (3) These interventions need numbers to make sense in topics where numbers don't intervene necessarily.
\subsection*{Class IX - 07.04}
\noindent \textbf{Part (I): Quantification} \\
\noindent Assigning numbers to things that don't naturally include numbers. Counting apples is not the same thing as counting utility. Numbers do not have to be divine or perfect --- they have to exist, we need a number of acres to inherit or how much wheat we have to sell. Same as how much money we have to pay for something. \\
\\
\noindent \textbf{Example 1:} numbers are not invented, they are estimated. Like, how on earth can you measure how much it costs per year the noise pollution in terms of heart diseases? \\
\noindent \textbf{Example 2:} how to measure the engagement of students on a subject? \\
\noindent \textbf{Example 3:} rider ratings in Uber. A driver's rating is interpretable for everybody, but there are different conceptions about what a ``good driver'' is. \\
\noindent \textbf{Example 4:} people needed to be compensated after the Chernobyl disaster: cancer, dead cattle, destroyed houses. How do you measure the \emph{damage}?
\\
\noindent Numbers transfer the authority to an argument where numbers are not organically used. It becomes an immensely complicated game, since it is a purely subjective matter. Once numbers are attached to things, we can start making new claims about them. \\
\\
\noindent \textbf{Part (II): Audit Cultures} \\
\noindent More numbers imply moral traction in the key of improvement, accountability and self-responsibility. Now that numbers are included, morality has taken place. \\
\\
\noindent Exams had no numbers back in the day: you either passed or not with oral examinations. Universities competed with each other, and Cambridge started taking written test. So a test grade became a number attached to a person. This was set so that the universities could figure out which one was the best. So people would strive to get a 10. You ought to be a ``10'' student: if you were a ``1'' student, it was not desirable. Numbers tell you what you \emph{should} be and what you are at the moment. A moral transfer from the number \emph{to} a person.\\
\\
\noindent This matters because it sets targets that you should aim for. The performance says a lot about someone in \emph{moral} terms. A low grade doesn't mean you are a \emph{bad} person, but a \emph{bad student}. \\
\\
\noindent \textbf{Example (I): Consultants' Billability.} Breeds a moral competition between workers because it measures their ``quality'', and their father company is not paying their salary but another company. This morality says something about your quality as a professional, and people \emph{aim} to a certain percent of billability. \\
\noindent \textbf{Example (II) ``Employee Engagement Score'':} it is a measure of employees' sentiments, based in a total shareholder return annualised. \\
\\
\noindent Numbers have the advantage that they travel from context to context. Nothing else can do this. Also, they work as a tool for detachment.
\subsection*{Class X - 09.04 (Consulting Cultures: The Beginnings)}
\noindent \textbf{Consulting} consists of providing a third party expertise for a fee. Today it is synonymous with business advisory. Why have most areas of the government been taken over by consulting companies? \\

\noindent \textbf{Types of consulting:} management consulting, technology/AI consulting, political consulting, healthcare consulting, government consulting. \\

\noindent \textbf{How consulting became ``a thing'':} In the early 1800s in the USA, surveys gave work to geologists, chemists, zoologists, and mineralogists. These ``maverick experts'' were called \emph{assayers}. They were almost exclusively single young men — geographically mobile, shrewd, self-promoters with a frontier spirit. Sons of merchants, farmers, and lawyers; little formal education but vast machine experience. They were not particularly well paid; consulting was something to do between contracts.

\begin{itemize}
	\item They worked across the West, the Alleghenies, the Appalachians, and upstate New York — anywhere capitalists prospected for ore or minerals.
	\item By the 1850s, this work called ``consulting'' was commonplace.
	\item Example: \textbf{Benjamin Silliman Jr.} — \emph{Report on Rock Oil / Petroleum} (1855). He became president of the Pennsylvania Rock Oil Company and the Bristol Mining Company, raising a clear conflict of interest.
\end{itemize}

\noindent \textbf{What did they do?} Surveys, material and sample testing, engineering experiments, machine and infrastructure development. They were hired by capitalists to decide whether to purchase lands, dig mines, design equipment, improve manufacturing, and market new products. Over time, consultants began expanding their jurisdiction from practical hands-on matters toward the managerial organisation of businesses. \\

\noindent \textbf{First management consulting firm:} 1886 — Arthur Dehon Little \& Roger B. Griffin, 103 Milk Street, Boston. \\

\noindent \textbf{Spirit of the time:} Frederick Winslow Taylor, \emph{Principles of Scientific Management} (1919): systematic management and managerial guidance would solve inefficiency. Arthur D. Little promised ``the perfection of products'' with ``the improvement of processes''. Management became something to reflect on through numbers, models, and abstractions — \emph{``a kind of abstract gaze onto company life''}. \\

\noindent \textbf{1890s--1920s:} Management consulting rose alongside US corporate capitalism. Firms went from small, owner-run, single-product operations to large, dispersed, decentralised corporations. Administration and management were increasingly separated from production — not just physically but logically and conceptually. \\

\noindent This growing division created ``administrative experts'': the management of different things (coal, postal service, farmland) became \emph{``just management''} — and consultants grew increasingly confident they could improve managerial tasks regardless of the specific activity. \\

\noindent \textbf{1930s:} The Great Depression consolidated American management consultancies. Many corporations faced bankruptcy and needed to drastically cut costs. Management consultancies stepped in, \emph{legitimising lay-offs} and allowing management to share the blame for unpopular decisions.

\subsection*{Class XI - 14.04 (Consultants: the work of producing abstract work)}
\noindent Source: Felix Stein, \emph{Work, Sleep, Repeat: The Abstract Labour of German Management Consultants}. \\

\noindent \textbf{Training camp anecdote} — Rahul (Excel coach): \emph{``All that remains after a client has paid millions of dollars for a [consulting] project is a deck [of printed PowerPoint slides], about this big.''} Training instilled speed and purposive use of time; consultants were systematically given far too much material to digest in too little time. Roleplay: the trainer pretended to be a distracted CEO, and consultants had to present results in a few seconds in an imaginary elevator or taxi. \\

\noindent \emph{``Being a management consultant combined a strong belief in our own importance with a considerable degree of cluelessness as to what we actually did: we mostly created Excels and Microsoft PowerPoint slides.''} \\

\noindent \textbf{Salaries:} EUR 60,000/year for the most junior consultants up to EUR 1,000,000+ for profit-sharing senior partners. \\

\noindent \textbf{What do management consultants actually do? They sustain, produce, and reproduce abstract work:}

\begin{enumerate}
	\item \textbf{Intellectual separation:} They produced and sold abstractions — Excels, PowerPoints, spreadsheets, ``management'', ``impact'', ``efficiency'' — far removed from anything concrete.
	\item \textbf{Separation from people:} Their work changed the work of people they never met. It was often possible to complete a project without spending any time with low-level employees.
	\item \textbf{Separation from processes:} They were several steps removed from actual economic value creation; they neither made nor sold the goods and services around which client companies were organised.
	\item \textbf{Separation from concreteness:} Consulting always remained opaque — for client companies and for the consultants themselves — filled with moments of ambiguity, opacity, and uncertainty.
\end{enumerate}

\noindent \textbf{Who were they?} Recent graduates from top universities — not because of skill, intelligence, or experience, but because they were ambitious, respected hierarchy, and could be trained into working long hours toward clearly defined deadlines (rites of passage — see Class III). \\

\noindent \textbf{Paper mill project (Stein):} The task was to look at the mill's work processes and pinpoint ``growth potential''. Two weeks; the consultants knew nothing about paper or paper mills. Their boss explained that the mill's management expected them to find at least EUR 30 million of ``potential'', with 50\% to come from labour costs.

\begin{itemize}
	\item Daily morning meetings and evening debriefs summarising ``improvement potential found''; minimised breaks; worked after dinner until late; slept approximately six hours per night.
	\item Method: examining forklift drivers — toured the mill, interviewed managers, observed routines, noted how enormous paper reels moved from the production line to storage, when and how often trucks came in, and how long it took to load them.
	\item Forklift activities were colour-coded: \textbf{``value added''} (green), \textbf{``enabler''} (yellow), \textbf{``non-value added''} (red). Assumption: if red activities could be cut out (by retraining staff or rearranging warehouses), the same amount of work could be done with fewer staff — yielding savings.
	\item The difference between ``what was'' and ``potential'' was expressed as a percentage change, then translated into people and salaries = \emph{``improvement potential''} (profit). Example: \emph{``warehouses 2 and 3 show potential of over 51 per cent''}.
	\item Similar analyses were applied throughout the mill: R\&D labs, raw material storage, interactions with suppliers and transport. Recommendations on how to speed up work processes were converted into potential monetary gains.
\end{itemize}

\noindent \textbf{Lucien} (French consultant): writing a PowerPoint slide for selling private travel insurance in Asia over 5 years. \emph{``I know nothing about insurance; absolutely nothing''} — yet managed to convince others of his expertise. \\

\noindent \textbf{Key question:} How can people who know nothing about a topic write plans involving millions of euros? Do clients not spot their ignorance? Are clients using consultants simply to enforce decisions that have already been made? \\

\noindent \textbf{Management consultants provide legitimacy} — because of both what they know and what they don't know. \\

\noindent \textbf{Why legitimacy?} Actual managers have little power; corporations are hugely bureaucratised and technocratised. Managers today tend to be equals (in education, work pathways, lifestyle) to those they manage, giving them little legitimacy over them. \emph{``[Consultants are] troops that you can send in somewhere. A little bit like a rapid deployment force, when you expect resistance and do not want to burn yourself or your own men.''} \\

\noindent \textbf{Legitimacy because of what they DON'T know:} Outsiders not bound up in social relations (loyalties, grudges). Even if the information needed existed somewhere in the company, it would take an outsider to judge its truthfulness and assess its importance. Consultants brought knowledge from other companies; having worked with competitors, their insights were transferable (anonymised) — with PowerPoints already organised per industry. \\

\noindent \textbf{Legitimacy because of what they DO know:} They were assigned offices inside the client company, given email accounts, employee cards, and canteen access; they had meetings with all heads of sections and middle management, and handled privileged and confidential information. \\

\noindent \textbf{Conclusion:} Consultants are hired as \emph{technologies of detachment} that shore up legitimacy for a decision, a project, or a line of thinking through their abstractions. Very often decisions had already been made and simply needed a legitimacy boost, or they were hanging in the air until someone had the courage to act. \\

\noindent Munich interviewee: \emph{``We are often hired by the top so as to push through decisions that have already been taken or have already become clear... `but consulting firm XY said we should [do this]' so it will get done.''} \\

\noindent \textbf{German health insurance case:} The basic problem lay in the abstractions needed to think with. Who knew what hundreds of employees spread across Germany were \emph{actually} up to at any given time? Those thinking in terms of ``relative staff increases'' claimed their divisions had originally been much larger than assumed. Others, thinking in terms of ``workload'', held their divisions were much smaller than HR assumed. Lacking concreteness and besieged by competing interests, they agreed to \emph{produce a number they would agree on} — not an accurate number but an \emph{exact} number that did the work of settling expectations and impressions from daily work at the company.

\subsection*{Class XII - 16.04 (Consultants and rhetorical technologies: Microsoft Excel, PowerPoint and the like)}
\noindent Source: Kimberly Chong, \emph{Best Practice: Management Consulting and the Ethics of Financialization in China}; Felix Stein, \emph{Work, Sleep, Repeat}. \\

\noindent Final reports were PowerPoint presentations; even internal reporting was often done in PowerPoint. Consultants hardly ever used a word processor; some colleagues used pre-drawn slide template paper. \textbf{The use of PowerPoint was a major reason why consulting work remained considerably opaque.} \\

\noindent \textbf{The aesthetics of PowerPoint:} Slides should \emph{appear to be} intuitive to the reader; they should elicit an almost automatic reaction — an effect of intuitive understanding. Aesthetics are critically important: consultants could spend several hours a day aligning texts, boxes, lines, and arrows, changing bullet point styles and font sizes, using round symbols instead of square ones, and ensuring high-quality printouts. The only complexity they allow is mathematical; they provide little context as to where content came from. Good consultants know how to direct attention towards graphs and numbers rather than background stories that would reveal the mess, disagreement, and complexities behind producing these abstractions. \\

\noindent \textbf{Narrative structure:} Presentations were written along a strict linear narrative structure — every slide logically building on the preceding one. Predefined story arcs: \textbf{situation, complication, solution}. A common technique was the \emph{``Mickey Mouse outline''} — minislides sketched on paper flip charts, looking like a comic strip, allowing the consultant to check the flow of a presentation's narrative and graphics. \\

\noindent Discussions of company goals came with pictures of a little arrow hitting a target; outstanding questions with a photo of a businessman scratching his head; \emph{next steps} with an image of footprints in the sand. This childishness stood in stark contrast to the presumed importance of the work. \\

\noindent \textbf{Bullet points as the basic unit of expression:} Short expressions often styled as headlines, preceded by a bullet sign, usually connected in a list. ``Thinking in bullet points'' is a cognitive style premised on list-making — rather than narrative or intellectual argument — as its organising principle, favouring abstraction and buzzwords over detail and nuance.

\begin{itemize}
	\item What matters when communicating through bullet points: whether text can be grouped under a thematic heading, not the logical relations between points or their order of priority.
	\item Compared to conventional prose, it makes it easy for a group of speakers to come to agreement on a description of the world. \emph{``It is a language well-suited to an opacity that diffuses disagreement.''}
	\item This kind of communication steers away from disagreements on facts and logic in favour of a mutual exchange of statements and tokens of recognition. It ``liberates'' communication from requirements of logical consistency and precision in the use of concepts.
	\item It allows entirely diverging understandings (e.g. what does ``resilience'' mean when my job is at stake?) without being disturbed by these contradictions.
\end{itemize}

\noindent \textbf{The Change Matrix} (Chong): a complex visual showing organisational health states — \emph{High Performance}, \emph{On Track}, \emph{Unsustainable}, \emph{Off Track} — with terms such as ``Burning platform'', ``Downward spiral'', and ``Building momentum''. It is a key consulting artefact: it appears technical and precise while remaining sufficiently vague to accommodate diverging interpretations, allowing consultants to impose a shared vocabulary on a client organisation without forcing agreement on the underlying facts.

\end{document}
